Developmental trajectories of word recognition: a priming study in bilinguals

### 2023-11-13 manuscript/manuscript.qmd
Finish introduction, fix participant's LP, and finish discussion

Building a mental lexicon is a major achievement in the development of an infant: by storing representations of how familiar words sound and what they mean, an infant is able to make sense of their linguistic input. The foundations of an initial lexicon are in place before their first year of life [@tincoff1999some; @halle1994emergence; @vihman2004cross; @parise2012electrophysiological; @bergelson20126; @bergelson2015early]. This initial lexicon consists of only a few items; mainly words for people, interjections, body parts, and food [@tincoff2012six; @tardif2008baby], but it undergoes rapid growth during the second year of life [@ganger2004reexamining; @goldfield1990early; @bloom2002children; @mcmurray2007defusing]. According to parental reports, the average 15-month-old infant already understands more than 100 words, and by two years, they understand more than 400 words [@frank2021variability]. This accelerated lexical developmental is reflected in infants' trajectories of word recognition: infants recognise familiar words faster and more efficiently as they approach their second birthday [@fernald1998rapid; @fernald2001half; @hurtado2007spoken]. Despite being exposed to a more complex linguistic input, bilinguals show equivalent trajectories of word acquisition and word recognition to their monolingual peers' [@vihman2007onset; @legacy2018vocabulary; @hoff2012dual; @de2014bilingual; @pearson1994patterns; @byers-heinlein2023sometimes]. This is a remarkable deed for two reasons. First, bilingual infants receive a relative impoverished linguistic input in each of their languages compared to monolinguals [@costa2014does]. Second, they face a more complex referential context, since they often learn two labels for each referent, one in each language. The mechanisms that allow bilingual' trajectories of lexical developmental to keep up with monolinguals' are still unclear.

Previous studies have pointed to the interplay between bilingual' languages as a candidate facilitative mechanism behind bilingual lexical acquisition. In particular, cross-language similarity at the lexical level has been associated with larger vocabulary sizes. @floccia2018introduction collected vocabulary data from a sample of 367 bilingual children living in the United Kingdom, who were learning English and an additional language (out of a diverse pool of 13 languages). Using an edit distance-based metric of phonological similarity, the authors computed the average lexical similarity between the translation pairs of the English vocabulary checklist, and of each of the other languages. Children's vocabulary sizes in the additional language was positively associated with the amount of language similarity between their two languages. For instance, children learning English and Dutch (which share high lexical similarity) showed larger vocabulary sizes than those learning English and Mandarin (which share low lexical similarity). The authors suggested that the acquisition of words in the additional language might be facilitated by their cognate status (i.e., being phonologically similar to their translation equivalent). Larger vocabulary sizes might then be expected in bilinguals learning two languages sharing a high proportion of cognates. Recent studies have reported an earlier acquisition of cognate words [@mitchell2022cognates; @garcia-castro2023cognate], providing further evidence that cross-language similarity at the lexical level might boost vocabulary growth in bilinguals.

The facilitative effect of cross-language similarity is in line with the language non-selective account of bilingual lexical access. There is robust evidence suggesting that bilingual adults activate both languages in parallel---even during monolingual situations---which influences their performance in word recognition and production tasks [@marian1999; @groot1992determinants; @schwartz2007reading; @dijkstra1999recognition; @dufour1995matching; @schwartz2007reading; @dijkstra2010cross]. For instance @costa2000cognate presented Catalan-Spanish bilinguals with a series of pictures of familiar objects. For each object, participants were asked to name it in Catalan (their dominant language). Unbeknownst to participants, the authors manipulated the cognate status pictures' labels in Catalan and Spanish. In half of the trials, the labels associated to the pictures were cognates (i.e., phonologically similar translation equivalents, e.g., *gat*-*gat* [*cat*]) in Catalan and Spanish, whereas in the other half of the trials the labels were non-cognates (e.g., *taula*-*mesa* [*table*]). Participants named pictures faster in cognate trials than in non-cognate trials. Critically, Spanish monolinguals who completed the same task, naming the pictures in Spanish, showed equivalent naming times in both conditions. These results revealed that bilinguals activated their Spanish phonology, despite performing the naming task exclusively in Catalan: the visual recognition of the presented pictures led to the activation of its associated phonological forms in both languages in parallel, which influenced the subsequent the dynamics of word production.

Parallel activation has also been reported in he developing lexicon [@poarch2012cross; @bosma2020cognate; @singh2014one; @jardak2019labels; @von2019impact; @floccia2020translation]. For instance, @von2012language found evidence of cross-language phonological priming in twenty 21- to 42-months-old children learning German and English. In their experimental task, each trials begun with the auditory presentation of an English prime word embedded in a carrier phrase. Then, the target label was auditorily presented in German. Finally, the target and a distractor pictures were presented side-by-side. The authors created three conditions. In the phonological priming condition, the auditorily presented prime and target labels were phonologically related (e.g., *slide*-*Klide* [*dress*]). In the priming through translation condition, both labels were phonologically related through translation: the auditorily presented prime label did *not* overlap with the target label (e.g., *leg*-*Stein* [*stone*]), but its translation in German did [*Bein*]. In the unrelated condition, prime and target labels were phonologically unrelated in both languages. Interestingly, participants showed a weaker target preference in priming through translation trials, compared to unrelated trials. This was interpreted evidence of the lexical representations of the prime labels being activated in both English and German, leading to interference effects in later target recognition.

Previous studies had reported similar cross-language priming effects. The conclusion that such effects occurred at the lexical level, as opposed to being the result of mere acoustic similarity, was nonetheless unwarranted. In studies where the auditorily presented prime label was phonologically overlapped with the target label [e.g., @marian1999activation], priming effects might be explained by the activation of the phonological segments of the prime label, which facilitated the recognition of the target label, which shared such phonological segments. This might have been the phonological priming trials in @von2012language, in which prime and target labels shared overt phonological overlap. In contrast, the priming through translation results suggest that cross-language activation in the developing lexicon occurs at the lexical level. Since participants' target recognition was modulated by the non-presented prime translation, such an inhibitory effect must stem at the lexical level. The fact that phonological priming facilitated the recognition of the target words (both labels belong to different languages), and that phonological priming through translation interfered target recognition (translation and target belong to the same language) points to cross-language connections playing an excitatory role, and to within-language connections playing an inhibitory role.

Despite extending the available knowledge on the developing bilingual lexicon, the study by @von2012language has some methodological challenges that prevent strong conclusions about language non-selective access. First, participants' received their English input at school, where they heard German and English (between three and four hours each) from a German native speaker. Except for four children who also listened to English at home, participants' only exposure to English was 3-4 hours per day, for the time they attended daycare (an average of 12.35 months). Assuming that infants are exposed to a daily average of 12 total hours of speech, and they attended daycare only during weekdays, their estimated exposure to English throughout the week represented between `r scales::percent((3*5*4*12.35)/(12*7*4*21), accuracy = 0.01)` and `r scales::percent((4*5*4*12.35)/(12*7*4*21), accuracy = 0.01)` for the 21 month-olds (the youngest in the sample), and between `r scales::percent((3*5*4*12.35)/(12*7*4*43))` and `r scales::percent((4*5*4*12.35)/(12*7*4*43))` for the 43-month-olds. Conventionally, children are considered bilingual if exposed to 25% or more of the time to a second language [@byers-heinlein2021multilab; @rocha-hidalgo2023defining]. Participants in @von2012language might therefore not be generalisable to other populations of bilinguals who received a more balanced dual language input. Second, the presence of cognates in the stimuli lists lead to some primes in the phonological priming condition sharing substantial cross-language phonological overlap thought translation with the target words like *fire* /\textipa{faIe}/--*Feuer* /\textipa{fOyeR}/ and *Eier* /\textipa{aIeR}/, or *Moon* /\textipa{mu:n}/--*Mond* /\textipa{mo:nd}/ and *Huhn* /\textipa{hu:n}/. Conversely, some trials in the phonological priming through translation shared within-language overlap with the target, such as *Soup* /\text{su:p}/--*Suppe* /\textipa{zupe}/---*Puppe* /\text{pupE}/ and *Puppe* /\textipa{pupE}/. Given that the number of total trials in both conditions amounted to six trials, a considerable proportion of the trials presented this methodological limitation. Finally, the observed pattern of results was dependent on the inclusion of a single item, *Buch* [*book*, in German]. When this item was included, monolinguals showed a similar interference effect to bilinguals in the priming through translation condition, in which only bilinguals with knowledge of the translation should show an effect. Additionally, given that monolinguals and bilinguals were not directly, it is not possible to conclude that the differences in statistical significance between monolingual and bilingual target looking, compared to chance levels, does not necessarily provide a significant difference between the two groups. As a consequence, the conclusion that the interference effects observed in the priming through translation condition is exclusive to bilinguals (which would suggest that the priming effect were not due to the intrinsic properties of the stimuli) in unwarranted.

Additionally, although the priming through translation paradigm provides a convenient framework to investigate parallel activation in the bilingual lexicon, its design prevents strong interpretations about the extent to which parallel activation is the result of the experimental manipulations of interest. The presentation of primes and targets in both languages in the same task---or even within the same trial---presents participants with a bilingual context that might lead to a higher baseline activation of lexical representations of both languages, contrary to a context in which parallel activation results from genuine cross-language activation [@grosjean1997bilingual]. Another limitation of priming through translation studies is that, when prime words are auditorily presented, the task constrains the naming context to one of the languages. The consequence is a potential asymmetry between the activation dynamics of the lexical representation of both labels: while the participant has heard one label, the activation of its translation must spread across semantic links. This might lead to a lower strength, and higher latency of the activation of the translation. In summary, current evidence on the cross-language interplay in the initial lexicon should be considered carefully.

In the present study, we put the hypothesis that bilingual infants activate both languages in parallel to a more severe test. We used an extension of the implicit naming task by @mani2010infant, in which participants were primed with a picture presented in silence. Previous studies using this paradigm have successfully provided evidence of language non-selectivity. In particular, @von2014bilinguals found that bilingual adults generate implicit labels in both labels for visually fixated pictures. The authors presented German-English bilinguals with 120 prime-target pairs. Primes were presented as familiar pictures in silence. After prime picture offset, target words were presented auditorily. Participants' N400 ERP components were recorded from target word onset. The authors manipulated the phonological relationship between the prime and target word-forms within participants' L1 and across L1 and L2. When prime and target were phonologically identical (*Affe*-*Affe*, German for *monkey*), or similar (*Fahne*-*Sahne*, German for *flag* and *ice-cream*, respectively) within L1, participants' showed a reduced N400 amplitude, compared to when prime and target were phonologically unrelated (*Messer*-*Seil*, German for *knife* and *rope*). Critically, a similar effect was found when prime and target were phonologically related through translation (*Rustsche*-*Kleid*, German for *slide* and *dress*). This suggests that participants activated prime labels in both languages in parallel, and that both labels impacted the dynamics of target words recognition.

This paradigm provides participants with an exclusively monolingual context in which they perform the task in only one of their languages. If differences in task performance are found associated with covert manipulation of cross-linguistic properties of the stimuli, such differences may be more conclusively linked to the parallel activation of lexical representations in participants' lexicons. By using an implicit naming paradigm, the activation of the lexical representation of the prime may occur in both languages as the result of lexical retrieval after visual recognition of the prime picture, since participants are not presented with any auditory prime label. This contrasts with priming through-translation tasks, in which the prime label in one of the languages is directly activated by auditory word recognition, while the label of its translation is expected to be activated as the result of cascaded activation spreading through their semantic association. This may lead to an asymmetry in the time course of the activation of both labels, which might in turn affect the dynamics of phonological priming in subsequent word recognition.

In the present investigation, we presented 20- to 32-months-old toddlers with a cross-linguistic phonological priming paradigm. We manipulated both the phonological overlap between the prime word and target word, and the cognate status of the prime word, to test the hypothesis that bilingual toddlers implicitly name familiar pictures in both languages. If so, phonological priming effect should be larger when the prime label in both languages shares phonological overlap with the target words (*Related/Cognate* condition), compared to when only the label of the prime in one language overlaps phonologically with the target (*Related/Non-cognate*). We also also included a control *Unrelated* condition in which regardless of the cognate status of the prime word, neither of the prime labels in any language overlapped phonologically with the target word. To test cross-language priming effects, we registered participants' time course of target picture looking in the three conditions. We predicted within-language priming effects, revealed as differences in target looking between *Unrelated* and *Related/Non-cognate* trials, as previously reported [@von2012language, @mani2010infant, and @mani2011phonological]. We further predicted cross-language priming effects, revealed as differences in target looking between *Related/Non-cognate* and *Related/Cognate* trials. 
This effect should be stronger in bilinguals, if not completely absent in monolinguals, who should be indifferent to the phonological properties of the stimuli in the other language.

Following previous studies associating the emergence of phonological links between lexical items to the size of children's vocabulary [@mani2011phonological], we also compared the explanatory power of two indices of receptive vocabulary size (dominant-language vocabulary size and total vocabulary size) against participants' age. We used parental reports of vocabulary checklist to estimate participants' vocabulary size. We computed vocabulary size in the dominant language (L1 vocabulary) as the proportion of words in the dominant language reported by their caregivers to be understood by the child in a vocabulary checklist, and the total vocabulary size as the proportion of words reported as understood by the child in *both* languages. If within-language or cross-language phonological associations in the lexicon emerge as a function of vocabulary growth, the size of the differences between conditions should larger in participants with larger vocabulary sizes. A model including participants' L1 or total vocabulary size as a predictor should then show better performance than a model that instead includes participants age. In the case of bilinguals, if the emergence of cross-language priming effects is associated with the growth of their vocabulary is *both* languages, a model including total vocabulary size as a predictor should perform better than a model that includes L1 vocabulary size. The opposite pattern would indicate that cross-language priming effects emerge as a function of the growth of L1 vocabulary.

@@

```{r tbl-participants}
#| label: tbl-participants
#| tbl-cap: "Demographic and linguistic profile of testing sessions."
participants_bcn |> 
	unnest_wider(doe) |> 
	select(-matches("others")) |> 
	arrange(lp, test_language) |> 
	summarise(n = n(),
			  n_participants = n_distinct(child_id),
			  across(age, lst(mean, sd, min, max)),
			  across(matches("doe_"), lst(mean, sd)),
			  # age_vctr = list(age),
			  .by = c(lp, test_language)) |> 
	gt(rowname_col = "test_language",
	   groupname_col = "lp") |> 
	# gt_plt_dist(age_vctr, type = "histogram") |>
	fmt_number(is.numeric, decimals = 1) |> 
	fmt_integer(c(n, n_participants)) |> 
	fmt_number(matches("doe"), decimals = 1, scale_by = 100) |> 
	cols_merge(c(age_min, age_max), 
			   pattern = "{1} ({2})") |> 
	cols_merge(c(age_mean, age_sd),
			   pattern = "{1} ({2})") |> 
	cols_merge(c(doe_catalan_mean, doe_catalan_sd),
			   pattern = "{1} ({2})") |> 
	cols_merge(c(doe_english_mean, doe_english_sd),
			   pattern = "{1} ({2})") |> 
	cols_merge(c(doe_spanish_mean, doe_spanish_sd), 
			   pattern = "{1} ({2})") |> 
	tab_spanner("Age (months)", matches("age")) |> 
	tab_spanner("Catalan", matches("catalan")) |> 
	tab_spanner("English", matches("english")) |> 
	tab_spanner("Spanish", matches("spanish")) |> 
	tab_spanner("Degree of Exposure (%)", matches("doe")) |> 
	cols_label(n = "N sessions",
			   n_participants = "N participants",
			   age_mean = "M (SD)",
			   age_min = "Range",
			   # age_vctr = "Distribution",
			   doe_catalan_mean = "M (SD)",
			   doe_english_mean = "M (SD)",
			   doe_spanish_mean = "M (SD)") |> 
	tab_style(cell_text(style = "italic"),
			  cells_column_labels()) |> 
	tab_style(cell_text(style = "italic"),
			  cells_column_spanners())

@@

We collected vocabulary data using parental responses to the Barcelona Vocabulary Inventory [BVQ, @garcia-castro2023bvq], an online vocabulary checklist inspired in several adaptations of the the Communicative Developmental Inventory [CDI, @fenson1994variability] developed to assess the vocabulary size of Catalan-Spanish bilingual toddlers. Families received a link to the BVQ immediately after each experimental session, and were given two weeks to fill it. We calculated two measures of receptive vocabulary size for each participant: *L1 vocabulary size* (proportion of words  reported as acquired in the checklist of the dominant language), and *total vocabulary size* (proportion of the words in both checklists reported as acquired). `r n_imputed` (`r prop_imputed`) Families failed to provide a complete response to the BVQ within the two-week time limit, or did not provide a successful response to the questionnaire. For missing questionnaire responses, we imputed the vocabulary size of the participant using single imputation, using the vocabulary size scores of a pool of `r n_pool` additional participants for which a successful response for the questionnaire had been gathered. We used participants age in months and their language profile (monolingual/bilingual) as predictors. We used the `mice` R package [@van2011mice] to perform imputation using the Bayesian linear regression method.

>>>>

Building a mental lexicon is a major achievement in the development of an infant: by storing representations of how familiar words sound and what they mean, an infant is able to make sense of their linguistic input. The foundations of an initial lexicon are in place before the end the first year of life [@tincoff1999some; @halle1994emergence; @vihman2004cross; @parise2012electrophysiological; @bergelson20126; @bergelson2015early]. This initial lexicon consists of only a few items; mainly words for people, interjections, body parts, and food [@tincoff2012six; @tardif2008baby], but it undergoes rapid growth during the second year of life [@ganger2004reexamining; @goldfield1990early; @bloom2002children; @mcmurray2007defusing; @bergelson2020comprehension]. According to parental reports, the average 15-month-old infant already understands more than 100 words, and by two years of age, they understand more than 400 [@frank2021variability]. This accelerated lexical developmental is reflected in infants' trajectories of word recognition: infants recognise familiar words faster and more efficiently as they approach their second birthday [@fernald1998rapid; @fernald2001half; @hurtado2007spoken]. Despite being exposed to a more complex linguistic input, bilinguals show equivalent trajectories of word acquisition and word recognition to their monolingual peers' [@vihman2007onset; @legacy2018vocabulary; @hoff2012dual; @de2014bilingual; @pearson1994patterns; @byers-heinlein2023sometimes; @bialystok2009bilingualism]. This is a remarkable deed for two reasons. First, bilingual infants receive a relative impoverished quantitative linguistic input in each of their languages, compared to monolinguals [@costa2014does; @cattani2014much; @thordardottir2011relationship]. Second, they face a more complex referential context, as they often learn two labels for each referent, one in each language [@au1990principle; @tsui2022translation; @bilson2015semantic; @de2006early]. The mechanisms that allow bilingual' trajectories of lexical developmental to keep up with monolinguals' are still unclear.

Previous studies have pointed to cross-language dynamics as a candidate mechanism boosting lexical acquisition in bilinguals [@floccia2018vocabulary; @blom2020cross; @gampe2021does]. In particular, lexical similarity between the two languages has been associated with larger vocabulary sizes in bilingual toddlers. @floccia2018vocabulary collected parental reports of vocabulary data from a sample of 367 bilingual children living in the United Kingdom, who were learning English and an additional language (out of a diverse pool of 13 languages). The authors then used an edit distance-based metric of phonological similarity to compute the average lexical similarity between the translation pairs in English and each of the additional languages. English and Dutch shared the highest lexical similarity, while English and Mandarin shared the lowest. Children's vocabulary sizes in the additional language was positively associated with the amount of language similarity between their two languages. For instance, children learning English and Dutch showed larger vocabulary sizes than those learning English and Mandarin. The authors suggested that the acquisition of words in the additional language might be facilitated by their cognate status (i.e., being phonologically similar to their translation equivalent). If this is the case, larger vocabulary sizes might then be expected in bilinguals learning two languages sharing a high proportion of cognates. This would be consistent with available evidence of an earlier acquisition of cognate words [@bosch2014first; @mitchell2022cognates; @garcia2023cognate].

The facilitative effect of cognateness is in line with the language non-selective account of bilingual lexical access. This account proposes that bilinguals activate both languages in parallel, even during monolingual situations, as suggested by robust experimental evidence [@marian1999activation; @groot1992determinants; @schwartz2007reading; @dijkstra1999recognition; @dufour1995matching; @dijkstra2010cross]. For instance @costa2000cognate presented Catalan-Spanish bilinguals with a series of pictures of familiar objects. Participants were asked to name each picture in Catalan, their dominant language. Unbeknownst to participants, the authors manipulated the cognate status pictures' labels in Catalan and Spanish. In half of the trials, the labels associated to the pictures were cognates (i.e., phonologically similar translation equivalents, e.g., *gat*-*gat* [*cat*]) in Catalan and Spanish, whereas in the other half of the trials the labels were non-cognates (e.g., *taula*-*mesa* [*table*]). Participants named pictures faster in cognate trials than in non-cognate trials. Critically, Spanish monolinguals---who completed the same task in Spanish---showed equivalent naming times in both conditions. These results revealed that bilinguals activated their Spanish phonology, despite performing the naming task exclusively in Catalan: the visual recognition of the presented pictures led to the parallel activation of its associated phonological forms in both languages, which influenced the subsequent dynamics of word production.

Parallel activation has also been reported in the developing lexicon [@poarch2012cross; @bosma2020cognate; @singh2014one; @jardak2019labels; @von2019impact; @floccia2020translation; @bosma2019longitudinal]. For instance, @von2012language found suggestive evidence of cross-language phonological priming in a sample of 20 German-English bilinguals aged 21 to 43 months. In their experimental task, each trial begun with the auditory presentation of an English prime word, followed by a target German label and a pair of target and distractor pictures. The authors registered participants' target picture looking (interpreted as a proxy of target word recognition) under three conditions. In the *phonological priming* condition, the auditorily presented prime and target labels were phonologically related (e.g., *slide*-*Klide* [*dress*]). In the *priming through translation* condition, prime and target were phonologically related through translation: the auditorily presented prime label did *not* overlap with the target label (e.g., *leg*-*Stein* [*stone*]), but its German translation did [*Bein*]. In the *unrelated* condition, prime and target labels were phonologically unrelated in both languages. Interestingly, participants showed a weaker target preference in priming through translation trials, compared to unrelated trials. This was interpreted evidence of the lexical representations of the prime labels being activated in both English and German, leading to interference effects in later target recognition. Previous studies had reported similar cross-language priming effects, but the conclusion that such effects occurred at the lexical level, as opposed to being the result of mere acoustic similarity, was nonetheless unwarranted. In studies where the auditorily presented prime label was phonologically overlapped with the target label [e.g., @marian1999activation], priming effects could be explained by sub-lexical facilitation across phonological segments. This might explain the facilitation effect found in *phonological priming* trials in @von2012language, in which English prime and German target labels shared overt phonological overlap. In contrast, the interference effect found in *priming through translation* trials points to cross-language competition at the lexical level. Since participants' target recognition was modulated by the phonology of the German translation of the prime (which participants did *not* hear), cross-language priming can be explained by the lexical co-activation of the English *and* German prime labels.

@von2012language extended the available knowledge on the developing bilingual lexicon, but some methodological challenges prevent drawing strong conclusions from their findings. First, bilingual participants' quantitative exposure to English as a second language was lower than what is conventionally considered as bilingualism [20-50% of cumulative exposure, @byers-heinlein2021multilab; @rocha-hidalgo2023defining]^[Except for four participants who were also exposed to English at home, participants in @von2012language received their English input exclusively at school from a German native speaker. Participants' exposure to each language ranged between three and four hours per day. This implies that participants' exposure to English consisted of an average of 3-4 hours per day, during the time they attended daycare (reported as an average 12.35 months). Assuming that infants were exposed to a daily average of 12 total hours of speech, and that they attended daycare five days per week, the estimated cumulative to English in the life of an average child in the sample was between `r scales::percent((3*5*4*12.35)/(12*7*4*21), accuracy = 0.01)` and `r scales::percent((4*5*4*12.35)/(12*7*4*21), accuracy = 0.01)` for the 21 month-olds (the youngest in the sample), and between `r scales::percent((3*5*4*12.35)/(12*7*4*43), accuracy = 0.01)` and `r scales::percent((4*5*4*12.35)/(12*7*4*43), accuracy = 0.01)` for the 43-month-olds. These percentages were calculated as:
$$\frac{\text{Total English input}}{ \text{Total English + German input}} = \frac{3 \text{ or } 4 \text{ hours/day} \times 5 \text{ days/week } \times 4\text{ weeks/month }  \times 21 \text{ or } 43 \text{ months of age}}{12 \text{ hours/day} \times 7 \text{ days/week } \times 4\text{ weeks/month }  \times 21 \text{ or } 43 \text{ months of age}}$$.]. Results should therefore be taken with caution when generalised to other populations of bilinguals. A second consideration is the presence of cognates in a substantial proportion of the trials in the *phonological priming* and *priming thorugh translation* conditions. Because of this, some prime labels in *phonological priming* trials shared substantial cross-language phonological overlap thought translation with the target words. Conversely, some trials in the *phonological priming through translation* shared within-language overlap with the target. Third, the observed pattern of results was dependent on the inclusion of a single item, *Buch* [*book*, in German]. When this item was included, monolinguals showed a similar interference effect to bilinguals' in the *priming through translation* condition, in which only bilinguals with knowledge of the translation should show an effect. Fourth, data from monolinguals and bilinguals were not compared directly (analyses were conducted separately for both groups). Differences in statistical significance between monolinguals and bilinguals across conditions do not necessarily guarantee significant differences between the two groups. As a consequence, whether the interference effects observed in the *priming through translation* condition is exclusive to bilinguals (which would suggest that the priming effect were not due to the intrinsic properties of the stimuli) or not is unclear. In summary, evidence of cross-language activation at the lexical level in infants is, to date, inconclusive.

In the present study, we use an extension of the implicit naming task [@mani2010infant; @mani2011phonological] to put the language non-selectivity in the developing lexicon to a more severe test. Previous studies have successfully used this task to investigate priming effects in word recognition in monolingual infants and bilingual adults. In the original implementation of the task, @mani2010infant primed infants with a picture presented in silence. At test, a target-distractor picture pair were presented side-by-side. Finally, the target picture's label was presented auditorily. The authors manipulated the phonological overlap between the prime and the target labels, so that in half of the trials both labels were phonologically related, sharing phonological onset (*cat*-*cup*), or phonologically unrelated (*ball*-*comb*). Prime and target-distractor pairs were semantically unrelated. Eighteen-month-old infants showed a stronger looking preference for the target picture after phonologically related prime pictures, compared to after phonologically unrelated primes. This suggests that infants implicitly named prime pictures despite such pictures having been presented in silence, and that the phonology of the resulting label interacted with the subsequent auditory recognition of the target label.

@von2014bilinguals adapted this paradigm to investigate cross-language activation in bilingual adults. The authors presented German-English bilinguals with 120 prime-target pairs. Primes were presented as familiar pictures in silence. After prime picture offset, target words were presented auditorily. Participants' eletrophysiological response was recorded, and N400 evoked potentials to the auditory target label was measured. The authors manipulated the phonological relationship between the prime and target word-forms within participants' L1 and across L1 and L2. When prime and target were phonologically identical (*Affe*-*Affe*, German for *monkey*), or similar (*Fahne*-*Sahne*, German for *flag* and *ice-cream*, respectively) within L1, participants' showed a reduced N400 amplitude, compared to when prime and target were phonologically unrelated (*Messer*-*Seil*, German for *knife* and *rope*). Critically, a similar effect was found when prime and target were phonologically related through translation (*Rustsche*-*Kleid*, German for *slide* and *dress*). This suggests that participants activated prime labels in both languages in parallel, and that both labels impacted the dynamics of target words recognition. In summary, the implicit naming paradigm has been successful used in the past to investigate the emergence of phonological priming effects in monolingual infants, and to investigate cross-language priming effects in bilingual adults.

The implicit naming paradigm provides several advantages over the priming through translation task. In the priming through translation task, primes and targets are auditorily presented in both languages in the same trial. This presents participants with a bilingual context that might lead to a higher baseline activation of lexical representations of both languages, contrary to a context in which parallel activation results from genuine cross-language activation [@grosjean1997bilingual]. Also, when prime words are auditorily presented, the task imposes an asymmetry between the activation dynamics of the lexical representation of the prime in each language: while participants hear prime labels in one language, the activation of prime labels in the other must occur via cascaded activation across semantic links with their translations. This might lead to a lower strength, and higher latency of the activation of prime labels in one language, compared to the presented labels in the other. In contrast, the implicit naming paradigm provides participants with an exclusively monolingual context in which they perform the task in only one of their languages. At the same time, the silent presentation of prime pictures allows participants to activate prime labels in both languages in a truly parallel fashion. If differences in task performance are found associated with covert manipulation of cross-linguistic properties of prime labels, such differences may be more conclusively linked to the parallel activation of lexical representations in both languages.

In the present investigation, we manipulated both the phonological overlap between the prime word and target word and the cognate status of the prime word, to test the whether bilingual toddlers implicitly name familiar pictures in both languages. If this is the case, phonological priming effects should be larger when the prime label in both languages shares phonological overlap with the target words (*Related/Cognate* condition), compared to when only the label of the prime in one language overlaps phonologically with the target (*Related/Non-cognate*). We also also included a control *Unrelated* condition in which regardless of the cognate status of the prime word, neither of the prime labels in any language overlapped phonologically with the target word. We registered participants' target looking during target-distractor presentation, and that, if within-language priming effects take place, differences in target looking between *Unrelated* and *Related/Non-cognate* trials should be observed, as previously reported [@von2012language, @mani2010infant; @mani2011phonological]. We further predicted cross-language priming effects, revealed as differences in target looking between *Related/Non-cognate* and *Related/Cognate* trials. This effect should be stronger in bilinguals than monolinguals, if not completely absent in the latter, who should be indifferent to the phonological properties of the stimuli in the other language.

Following previous studies associating the emergence of phonological links between lexical items to the size of children's vocabulary [@mani2011phonological; @chow2017spoken], we also compared the explanatory power of two indices of receptive vocabulary size against participants' age: dominant-language vocabulary size and total vocabulary size. If within-language or cross-language phonological associations in the lexicon emerge as a function of vocabulary growth, the size of the differences between conditions should larger in participants with larger vocabulary sizes. A model including participants' dominant-language or total vocabulary size as a predictor should then show better performance than a model that instead includes participants age. In the case of bilinguals, if the emergence of cross-language priming effects is associated with the growth of their vocabulary is *both* languages, a model including total vocabulary size as a predictor should perform better than a model that includes dominant-language vocabulary size. The opposite pattern would indicate that cross-language priming effects emerge as a function of the growth of dominant-language vocabulary.

@@

```{r tbl-participants}
#| label: tbl-participants
#| tbl-cap: "Demographic and linguistic profile of testing sessions."
participants_bcn |> 
	unnest_wider(doe) |> 
	select(-matches("others")) |> 
	arrange(lp, test_language) |> 
	distinct(session_id, child_id, .keep_all = TRUE) |> 
	summarise(n = n(),
			  n_participants = length(unique(child_id)),
			  across(age, lst(mean, sd, min, max)),
			  across(matches("doe_"), lst(mean, sd)),
			  # age_vctr = list(age),
			  .by = c(lp, test_language)) |> 
	gt(rowname_col = "test_language",
	   groupname_col = "lp") |> 
	summary_rows(groups = c("Monolingual", "Bilingual"),
				 columns = matches("age"),
				 fns = list(Total ~ mean(.)),
				 fmt = ~ fmt_number(., decimals = 1, use_seps = FALSE)) |> 
	summary_rows(groups = c("Monolingual", "Bilingual"),
				 columns = c(n, n_participants),
				 fns = list(Total ~ sum(.)),
				 fmt = ~ fmt_integer(.)) |> 
	summary_rows(groups = c("Monolingual", "Bilingual"),
				 columns = matches("doe"),
				 fns = list(Total ~ mean(.)),
				 fmt = ~ fmt_number(., decimals = 1, scale_by = 100)) |> 
	grand_summary_rows(columns = matches("age"),
					   fns = list(Total ~ mean(.)),
					   fmt = ~ fmt_number(., decimals = 1)) |> 
	grand_summary_rows(columns = matches("doe"),
					   fns = list(Total ~ mean(.)),
					   fmt = ~ fmt_number(., decimals = 1, scale_by = 100)) |> 
	grand_summary_rows(columns = c(n, n_participants),
					   fns = list(Total ~ sum(.)),
					   fmt = ~ fmt_integer(.)) |> 
	# gt_plt_dist(age_vctr, type = "histogram") |>
	fmt_number(is.numeric, decimals = 1) |> 
	fmt_integer(c(n, n_participants)) |> 
	fmt_number(matches("doe"), decimals = 1, scale_by = 100) |> 
	cols_merge(c(age_min, age_max), 
			   pattern = "{1}-{2}") |> 
	cols_merge(c(age_mean, age_sd),
			   pattern = "{1} ({2})") |> 
	cols_merge(c(doe_catalan_mean, doe_catalan_sd),
			   pattern = "{1} ({2})") |> 
	cols_merge(c(doe_english_mean, doe_english_sd),
			   pattern = "{1} ({2})") |> 
	cols_merge(c(doe_spanish_mean, doe_spanish_sd), 
			   pattern = "{1} ({2})") |> 
	tab_spanner("Age (months)", matches("age")) |> 
	tab_spanner("Catalan", matches("catalan")) |> 
	tab_spanner("English", matches("english")) |> 
	tab_spanner("Spanish", matches("spanish")) |> 
	tab_spanner("Degree of Exposure (%)", matches("doe")) |> 
	cols_label(n = "Sessions",
			   n_participants = "Participants",
			   age_mean = "M (SD)",
			   age_min = "Range",
			   doe_catalan_mean = "M (SD)",
			   doe_english_mean = "M (SD)",
			   doe_spanish_mean = "M (SD)") |> 
	tab_style(cell_text(style = "italic"),
			  cells_column_labels()) |> 
	tab_style(cell_text(style = "italic"),
			  cells_column_spanners())

@@

We collected vocabulary data using parental responses to the Barcelona Vocabulary Inventory [BVQ, @garcia-castro2023bvq], an online vocabulary checklist developed to assess the vocabulary size of Catalan-Spanish bilingual toddlers, and inspired in several adaptations of the the Communicative Developmental Inventory [CDI, @fenson1994variability]. Families received a link to the BVQ immediately after each experimental session, and were given two weeks to fill it. We calculated two measures of receptive vocabulary size for each participant: *L1 vocabulary size* (proportion of words  reported as acquired in the checklist of the dominant language), and *total vocabulary size* (proportion of the words in both checklists reported as acquired). `r english::Words(n_imputed)` (`r prop_imputed`) families failed to provide a complete response to the BVQ within the two-week time limit. We imputed missing vocabulary size scores using single imputation, taking the vocabulary size scores of a pool of `r n_pool` additional participants for which a successful response for the questionnaire had been gathered. We used participants' age in months and their language profile (monolingual or bilingual) as predictors. We used the `mice` R package [@van2011mice] to perform imputation using the Bayesian linear regression method (see Appendix B).
